### 2015-10-29 manuscript.Rnw
jay-double-ewe-bee edits

Advances in sequencing and computing technologies have lead to a revolution in
systematic biology. The ability to routinely generate molecular datasets from
any extant organism has allowed researchers to resolve long-running taxonomic
disputes and estimate phylogenies for previously understudied groups. In
parallel, the ease with which phylogenies can be estimated has spurred the
development of novel phylogenetic comparative methods. These methods allow
researchers to explore fundamental questions about the origin of biodiversity,
including the evolution of morphological and ecological traits, the
spatio-temporal variation in speciation rates, or both
\citep{OMeara2012,Pennell2013}.

Ideally, the ever-increasing number of published phylogenies would contribute to
a synthesis of phylogenetic knowledge, ultimately leading to a richer understanding of
the history of life, while at the same time providing high-quality phylogenetic information for use
in comparative analyses. However, in practice synthesizing phylogenetic data
is a difficult task. Phylogenetic information is scattered, often only available as image
files within publications, and the lack of standardization to store and
represent phylogenetic data makes it difficult for researchers to access,
synthesize, and integrate this information into their own research
(\citealt{Stoltzfus2012}, \citealt{Drew2013}; but see \citealt{Cranston2014} for suggestions of best
practices).

The Open Tree of Life (OTL) project aims at assembling and synthesizing our
current understanding of phylogenetic relationships across all organisms on
Earth \citep{Hinchliff2015}, while providing tools and services that facilitate
access to this information. OTL combines taxonomic information that serves as
the backbone for the phylogenetic relationships, and published phylogenies to
elucidate relationships among taxa. This combination of information is used to
form a comprehensive synthetic tree. Studies can be contributed to the synthetic tree
through a curator interface (\url{https://tree.opentreeoflife.org/curator}),
allowing the synthetic tree to be continuously updated as relationships are
elucidated or reevaluated. The current draft of the OTL synthetic tree contains
2.3 million tips. Beyond obvious applications across the life sciences to
explore questions in evolution, biodiversity, and conservation, the resources
OTL provides is useful for education and outreach (e.g., illustrating course
material, or developing outreach activities to explore relationships among species).

@@

The OTL project provides a web interface to explore, query and
download parts of the synthetic tree (\url{http://tree.opentreeoflife.org})
as well as the published trees that contribute to it. Additionally, the project
provides Application Programming Interfaces (APIs)  to access programmatically
the different services and underlying data.  Here, we present \texttt{rotl}, a
package to interact with OTL's APIs to search for and import phylogenetic trees
and taxonomic information directly in R. Having access to these data in R allows
researchers to take advantage of the multiple packages that are available to
visualize, analyze, and manipulate phylogenetic and comparative data (e.g., ape
\citealt{Paradis2004}, phytools \citeateu{Revell2012}, geiger
\citealt{Pennell2014-geiger}, ggtree \citealt{Yu-ggtree}, phylobase
\citealt{phylobase-0.8.0}, RNeXML \citealt{Boettiger2015}; see
\url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
comprehensive list). There is also a growing list of packages that allow users
to query and access data from the web directly in R: \hl{rFISHBASE
10.1111/j.1095-8649.2012.03464.x, paleoDB 10.1111/ecog.01154, rAvis
10.1371/journal.pone.0091650 \dots}. By providing direct access to high quality
phylogenetic data in R, \texttt{rotl} fills a key gap in typical comparative
analysis workflows. Moreover, by allowing the gathering, processing, and analysis
of data to be performed programmatically, \texttt{rotl} facilitates reproducible
research practices.

@@

\begin{enumerate}
\item The \emph{taxonomy} used as the backbone of the tree, the Open Tree
  Taxonomy (OTT);
\item The \emph{studies} and their associated trees, some of which are chosen by
  curators to assemble the synthetic tree;
\item A \emph{taxonomic name resolution service} (TNRS) used to match taxon
  names to the Open Tree Taxonomy identifiers;
\item The \emph{synthetic tree} itself, the `Open Tree'.
\end{enumerate}

Additionally, the project provides access to other data, the Graph of Life, that
stores information about the underlying tree alignment graphs \citep{Smith2013}
used to synthesize the various sources of information from which the Open Tree
is assembled. Data from this API are not intended for normal users, and its
use will not be demonstrated in this paper, although \texttt{rotl} has functions to
retrieve data it provides.

\texttt{rotl} gives users access to all endpoints provided by version 2 of the
OTL APIs. Phylogenetic trees served by the API can be imported directly into
R's memory and are represented using the \texttt{ape} \citep{Paradis2004} tree
structure (objects of class \texttt{phylo}), or can be written to files in the
Newick, NEXUS \citep{Maddison1997} or NeXML \citep{Vos2012} file formats. This
allows researchers to use these trees either directly with other R packages, or
to be imported in other programs that make use of phylogenetic tree files.

@@

% Tree versionning? Replicability of analyses

@@

The package is well-documented, and includes two package
vignettes (documents that demonstrate the use of the package and contain executable 
R code)
% as well as per-function documentation
. There is also an extensive
test-suite that covers both the internal functions that \texttt{rotl} uses to
connect to OTL and the public functions users apply to access and process data.

@@

To illustrate how to obtain relationships from a set of taxa, we will rely on
OTL to draw a phylogenetic tree for a set of model organisms.
First, we use the function \texttt{tnrs\_match\_names} to match the taxon 
names to their Open Taxonomy identifiers.
<<init, echo=FALSE, purl=FALSE>>=
library(rotl)
@

%First, we use the function \texttt{tnrs\_match\_names} to match the taxon names to their Open Taxonomy identifiers.

<<get_taxa, cache=TRUE, purl=FALSE>>=
taxa <- tnrs_match_names(names = c("Escherichia colli",
                                   "Chlamydomonas reinhardtii",
                                   "Drosophila melanogaster",
                                   "Arabidopsis thaliana",
                                   "Rattus norvegicus",
                                   "Mus musculus",
                                   "Cavia porcellus",
                                   "Xenopus laevis",
                                   "Saccharomyces cervisae",
                                   "Danio rerio"))
@

@@

<<plot_taxa, fig.height=4, cache=TRUE, purl=FALSE, fig.cap="The phylogenetic tree returned by OTL for the list of model species used as an example.">>=
tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]])
plot(tree, cex=.8, label.offset=.1, no.margin=TRUE)
@

@@

<<plot_cats, purl=FALSE, fig.height=4.2, echo=FALSE, fig.cap='Phylogeny of the Felidae published in \\citealt{Johnson2006} and retrieved from OTL using \\texttt{rotl}'>>=
plot(cat_tree, label.offset=.0001, cex=.75, no.margin=TRUE)
@

@@

In recent years, R has become an essential part of the toolbox of many researchers in
evolutionary biology. R greatly facilitates the analysis of large datasets, and
allows researchers to combine methods in novel ways because many methods for
comparative analyses are implemented, and because it is a relatively easy to use
programming language. Additionally, as more data are made available online and
accessible using APIs, several packages have been developed to interact and
download these datasets directly in R, enabling direct and reproducible analyses.
Notably, the organization rOpenSci (\url{https://ropensci.org}) has fostered a
community of researchers who develop tools and methods to facilitate the use of
open data, as well as broaden the adoption of open science practices in
general. \texttt{rotl} contributes to this initiative, and data from OTL can be
combined with other sources easily. For instance in Figure~\ref{fig:cat_map}, we
show how we can obtain a map of the occurrences for some of the cat species
(genus \textit{Lynx}) included in the phylogeny above using records for these
species found in GBIF (code available in the Appendix). Another example is available
in the package vignette ``Data mashups''.

<<cat_map, purl=TRUE, echo=FALSE, size='small', fig.height=4.2, results='hide', message=FALSE, cache=TRUE, fig.cap='GBIF records for the species in \\textit{Lynx} included in the phylogeny associated with the study by \\citealt{Johnson2006}'>>=
## load required packages
library(rotl)
library(rgbif)
library(ggplot2)
library(maps)

## Import the felid tree using study and tree IDs discovered with 
## studies_find_studies() in the manuscript

cat_tree <- get_study_tree(study_id ="pg_1981",                           
                           tree_id = "tree4052")

@@

\texttt{rotl} is free, open source, and released under a Simplified
BSD license. Stable versions are available from the CRAN repository
(\url{https://cran.r-project.org/package=rotl}), and development
versions are available from GitHub
(\url{https://github.com/ropensci/rotl}). The package is under active development,
and authors welcome bug reports of feature requests via the github repository.
The source for this manuscript is available at \url{https://github.com/fmichonneau/rotl-ms}.

@@

We would like to thank the organizers of the OpenTree of Life APIs hackathon
that was held at the University of Michigan, Ann Arbor, 15-19 September, 2014,
where the development of \texttt{rotl} was started. We would also like to thank
Scott Chamberlain (rOpenSci) for providing a thorough code review. FM was
supported by iDigBio, and therefore this material is based upon work supported
by the National Science Foundation’s Advancing Digitization of Biodiversity
Collections Program (Cooperative Agreement EF-1115210)

>>>>

Advances in sequencing and computing technologies have lead to a revolution in
systematic biology. The ability to routinely generate molecular datasets from
any extant organism has allowed researchers to resolve long-running taxonomic
disputes and estimate phylogenies for previously understudied groups. In
parallel, the ease with which phylogenies can be estimated has spurred the
development of new phylogenetic comparative methods. These methods allow
researchers to explore fundamental questions about the origin of biodiversity
including the evolution of morphological and ecological traits, the
spatio-temporal variation in speciation rates, or both
\citep{OMeara2012, Pennell2013}.

Ideally, the ever increasing number of published phylogenies would contribute to
a synthesis of phylogenetic knowledge, ultimately leading to a better understanding of
the history of life while at the same time providing high-quality phylogenetic information 
for use in comparative analyses. However, in practice synthesizing phylogenetic data
is a difficult task. Phylogenetic information is largely scattered, often only available as image
files within publications, and the lack of standardization to store and
represent phylogenetic data makes it difficult for researchers to access,
synthesize, and integrate this information into their own research
(\citealt{Stoltzfus2012}; \citealt{Drew2013}; but see \citealt{Cranston2014} for suggestions of best
practices).

The Open Tree of Life (OTL) project aims at assembling and synthesizing our
current understanding of phylogenetic relationships across all organisms on
Earth \citep{Hinchliff2015}, while providing tools and services that facilitate
access to this information. OTL combines taxonomic information that serves as
the backbone for the phylogenetic relationships, and published phylogenies to
elucidate relationships among taxa. This combination of information is used to
structure the comprehensive synthetic tree. Studies can be contributed to the synthetic 
tree through a curator interface (\url{https://tree.opentreeoflife.org/curator}),
allowing the synthetic tree to be continuously updated as relationships are
elucidated or reevaluated. The current draft of the OTL synthetic tree contains
2.3 million tips. Beyond obvious applications across the life sciences to
explore questions in evolution, biodiversity, and conservation, the resources
OTL provides is useful for education and outreach (e.g., illustrating course
material, or developing outreach activities to explore relationships among species).

@@

The R programming language is a popular tool for  phylogenetics and comparative
analysis. The R packages \texttt{ape} \citep{Paradis2004}, \texttt{phylobase}
\citep{phylobase-0.8.0}, and \texttt{RNeXML} \citep{Boettiger2015} each provide
functions to import and manipulate trees within R and save the results in
standard data formats. Additional packages including \texttt{phytools}
\citep{Revell2012}, \texttt{geiger} \citep{Pennell2014-geiger}, and \texttt{ggtree} 
\citep{Yu-ggtree} allow users to analyse and visualise data in a 
phylogenetic context (see 
\url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
comprehensive list of phylogenetics packages in R). In addition to packages for
phylogentic and comparative analyses, a growing number of R packages allow users to query and
access data from the web (e.g. \texttt{rFISHBASE} \citep{Boettiger2012}, \texttt{rAvis} \citep{Varela2014}, and \texttt{paleoDB} \citep{Varela2015}), meaning data 
associated with taxa in a given phylogeny can be obtained directly in R. 
These packages, combined with the language's support for literate programming 
(\citealt{Knuth1984}; \citealt{Xie2014}), make R an ideal platform for reproducible research
in phylogenetics and comparative biology, as they allow a complete record of the 
steps taken in gathering, processing, and analysing a given data set to be
produced.

Here we present \texttt{rotl}, and R package that allows users to download
phylogenetic and taxonomic data from the OTL directly in R. \texttt{rotl} takes
advantage of Application Programming Interfaces (APIs) provided by OTL to access
sub-trees from the synthetic Open Tree, as well as the published source trees that 
contribute to the synthesis. By providing direct access to high quality
phylogenetic data in R, \texttt{rotl} fills a key gap typical comparative
analysis workflows, and extends the degree to which R supports reproducable
research in phylogenetics and comparative biology.

@@

\begin{enumerate}
\item The \emph{taxonomy} used as the backbone of the tree, the Open Tree
  Taxonomy (OTT);
\item The \emph{studies} and their associated trees some of which are chosen by
  curators to assemble the synthetic tree;
\item A \emph{taxonomic name resolution service} (TNRS) used to match taxon
  names to the Open Tree Taxonomy identifiers;
\item The \emph{synthetic tree} itself, the `Open Tree'.
\end{enumerate}

Additionally, the project provides access to other data, the Graph of Life, that
stores information about the underlying tree alignment graph \citep{Smith2013}
used to synthesize the various sources of information from which the Open Tree
is assembled. Data from this API are not intended for normal users, and its
use will not be demonstrated in this paper, although \texttt{rotl} has functions to
retrieve data it provides.

\texttt{rotl} gives users access to all endpoints provided by version 2 of the
OTL APIs. Phylogenetic trees served by the API can be imported directly into
R's memory and are represented using the \texttt{ape} \citep{Paradis2004} tree
structure (objects of class \texttt{phylo}), or can be written to files in the
Newick, NEXUS \citep{Maddison1997}, or NeXML \citep{Vos2012} file formats. This
allows researchers to use these trees either directly with other R packages, or
to be imported in other programs that make use of phylogenetic tree files.

@@

% Tree versioning? Replicability of analyses

@@

The package is well-documented, and includes two package
vignettes (documents that demonstrate the use of the package and contain executable 
R code)
%as well as per-function documentation
. There is also an extensive
test-suite that covers both the internal functions that \texttt{rotl} uses to
connect to OTL and public functions users apply to acess and process data.

@@

To illustrate how to obtain relationships from a set of taxa, we will rely on
OTL to draw a phylogenetic tree for a set of model organisms. First, we use the function 
\texttt{tnrs\_match\_names} to match the taxon names to their Open Taxonomy identifiers.

<<init, echo=FALSE, purl=FALSE>>=
library(rotl)
@
<<get_taxa, cache=TRUE, purl=FALSE>>=
taxa <- tnrs_match_names(names = c("Escherichia colli",
                                   "Chlamydomonas reinhardtii",
                                   "Drosophila melanogaster",
                                   "Arabidopsis thaliana",
                                   "Rattus norvegicus",
                                   "Mus musculus",
                                   "Cavia porcellus",
                                   "Xenopus laevis",
                                   "Saccharomyces cervisae",
                                   "Danio rerio"))
@

@@

<<plot_taxa, fig.height=4, cache=TRUE, purl=FALSE, fig.cap="The phylogenetic tree returned by OTL for the list of model species used as an example.">>=
tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]])
plot(tree, cex = .8, label.offset = .1, no.margin = TRUE)
@

@@

<<plot_cats, purl=FALSE, fig.height=4.2, echo=FALSE, fig.cap='Phylogeny of the Felidae published in \\citealt{Johnson2006} and retrieved from OTL using \\texttt{rotl}.'>>=
plot(cat_tree, label.offset=.0001, cex=.75, no.margin=TRUE)
@

@@

In recent years, R has become an essential part of the toolbox of many researchers in
evolutionary biology. R greatly facilitates the analysis of large datasets, and
allows researchers to combine methods in novel ways because many methods for
comparative analyses are implemented, and because it is a relatively easy to use
programming language. Additionally, as more data are made available online and
accessible using APIs, several packages have been developed to interact and
download these datasets directly in R, thereby enabling direct and reproducible analyses.
Notably, the organization rOpenSci (\url{https://ropensci.org}) has fostered a
community of researchers who develop tools and methods to facilitate the use of
open data as well broaden the adoption of open science practices in
general. For instance, the rOpenSci-developed packge \texttt{TreeBase} \citep{Boettiger2012TB} allows
users to access phylogenies stored in treeBASE (\url{www.treebase.org}). 
\texttt{rotl} contributes to this
initiative, and greatly extends the number of taxa for which phylogenetic data
can be retrieved within R, while allowing the  data from OTL can be combined with 
other sources easily.

The package vignettes provide two demonstrations of the integration of a
phylogeny and data associated the taxa it represents. Specifically, the 
``Data mashups'' vignette provides an example of how data associated with tip-taxa
can be gathered and visualized. Another vignette titled ``meta-analysis'' 
demonstrates how a complete analysis, including the gathering of data and a
phylogeny, can be performed in a single R session. We further demonstrate the
integration of phylogenetic data retrieved using \texttt{rotl} with other 
data here. We show how we can obtain a map of the occurrences for some of the cat species
(genus \textit{Lynx}) included in the phylogeny above using records for these
%species found in GBIF (Figure~\ref{fig:cat_map}; code available in the Appendix).

<<cat_map, purl=TRUE, echo=FALSE, size='small', fig.height=4.2, results='hide', message=FALSE, cache=TRUE, fig.cap='GBIF records for the species in \\textit{Lynx} included in the phylogeny associated with the study by \\citealt{Johnson2006}.'>>=
## load required packages
library(rotl)
library(rgbif)
library(ggplot2)
library(maps)

## Import the felid tree using study and tree IDs discovered with 
## studies_find_studies() in the manuscript
cat_tree <- get_study_tree(study_id ="pg_1981",                           
                           tree_id = "tree4052")

@@

\texttt{rotl} is free, open source, and released under a Simplified
BSD license. Stable versions are available from the CRAN repository
(\url{https://cran.r-project.org/package=rotl}), and development
versions are available from GitHub
(\url{https://github.com/ropensci/rotl}). The package is under active development,
and authors welcome bug reports of feature requests via the github repository.
The source for this manuscript is available on GitHub (\url{https://github.com/fmichonneau/rotl-ms}).

@@

We would like to thank the organizers of the OpenTree of Life APIs hackathon
that was held at the University of Michigan, Ann Arbor, 15-19 September, 2014,
where the development of \texttt{rotl} was started. We would also like to thank
Scott Chamberlain (rOpenSci) for providing a thorough code review. FM was
supported by iDigBio, and therefore this material is based upon work supported
by the National Science Foundation’s Advancing Digitization of Biodiversity
Collections Program (Cooperative Agreement EF-1115210).
